---
title: "GCclassifier: an R package for the prediction of molecular subtypes of gastric cancer"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: TRUE
bibliography: bibliography.bib
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{GCclassifier: an R package for the prediction of molecular subtypes of gastric cancer}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 9,
  fig.align = 'center',
  collapse = TRUE,
  message = TRUE,
  echo = TRUE,
  warnings = FALSE
)
```

Jiang Li^1,2^, Xin Wang^1,2,3^

^1^ Department of Surgery, The Chinese University of Hong Kong, Hong Kong SAR, China.\
^2^ Department of Biomedical Sciences, City University of Hong Kong, Hong Kong SAR, China.\
^3^ Shenzhen Research Institute, The Chinese University of Hong Kong, Shenzhen, China.

-   Contact: [xinwang\@cuhk.edu.hk](mailto:xinwang@cuhk.edu.hk){.email}   
-   Package: `r BiocStyle::pkg_ver("GCclassifier")`

# Introduction

The `GCclassifier` package provides molecular subtype prediction of gastric cancer using log2 scaled gene expression profiles. The core classification algorithm is [random forest](https://www.stat.berkeley.edu/~breiman/RandomForests/). Moreover, besides R functions, it also provides local and online Shiny applications which facilitates the non-R users to make gastric cancer molecular subtype prediction. Moreover, the minimal R version required to install the `GCclassifier` package is 4.1.0.

# Input data

`GCclassifier` provides robust prediction performance on microarray & RNA-seq datasets. Gene IDs in gene expression profiles must be [NCBI Entrez](https://www.ncbi.nlm.nih.gov/gene), [Ensembl](http://www.ensembl.org/index.html), [HGNC symbol](http://www.genenames.org/) or [Refseq](https://www.ncbi.nlm.nih.gov/refseq/), and `colnames` should be set to sample names. More specifically:

-   Microarray data input should be pre-processed and normalized (log2 transformed).\
-   RNA-sequencing data input should be log2 transformed normalized values (log2 TPM recommended).\
-   At least two samples were required in the microarray/RNA-seq data.

If the input dataset is not log2 transformed, the function will detect the gene expression profiles range and perform the log2 transformation automatically if needed.

# Quick start

## Pre-requisite of installation

Before running the example codes in the vignette, the following R packages are required:

-   dplyr (dataframe manipulation)
-   magrittr (pipe operator)
-   randomForest (prediction classifier)
-   BiocManager (package installation)
-   shiny (Shiny webpage)
-   DT (interactive tables in Shiny)
-   shinyjs (shiny webpage)
-   impute (missing data imputation)
-   AnnotationDbi (gene identifier converter)
-   org.Hs.eg.db (gene identifier converter)

```{r, eval=FALSE}
# Required packages: run if not already installed
if(!requireNamespace('BiocManager')){
  install.packages('BiocManager')
}

BiocManager::install(c(
  'impute', 'dplyr', 'magrittr', 'AnnotationDbi', 'randomForest', 'org.Hs.eg.db', 'shiny', 'DT', 'shinyjs'
)， force = T)
```

## Load example dataset GSE62254

In this case study, we would use `GCclassifier` package to perform gastric cancer molecular subtype prediction on gene expression profiles. Specifically, this gene expression profiles (ACRG cohort) are from a microarray dataset of 300 gastric cancer patients named [GSE62254](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE62254) (@Cristescu2015). Using this gene expression profiles, we will predict the molecular subtype of both three subtype systems (`EMP` (@Oh2018), `ACRG`(@Cristescu2015) and `TCGA`(@TCGA2014)).

```{r, eval=TRUE}
  ## Load example data set from GCclassifier and have a glimpse
  library(GCclassifier)
  data("GSE62254")
  data("GSE62254_subtype")
  GSE62254[1:5, 1:5]
  
  ## Check whether sample names in expression profile is identical to the sample names in clinical information
  identical(colnames(GSE62254), GSE62254_subtype$GEO_ID)
```

The gene expression profiles were downloaded using R package `GEOquery`, and the expression levels were mapped from probes to gene symbols, and if multiple probes were mapped to the same gene, the probe with the maximum expression value would be adopted. In GSE62254 dataset, the `rownames(GSE62254)` were gene symbols and the `colnames(GSE62254)` were the sample names.

## Case study 1: The EMP subtype prediction

To predict the EMP subtypes, the gene expression profiles of GSE62254 would be the input data and more information of other parameters would be found in the help page of `classifyGC`:

```{r, eval=TRUE, message=TRUE}
  emp.res <- classifyGC(
    Expr = GSE62254, ## gene expression profile with log2 transformation
    method = 'EMP', ## subtyping system
    idType = 'SYMBOL', ## the gene identifier type in gene expression profile
    useMinPosterior = F ## whether the minPosterior threshold is also used for EMT subtyping
  )
  
  table(emp.res$subtype, GSE62254_subtype$Subgroup)
```
One thing to be noticed, the useMinPosterior parameter indicates whether the minPosterior threshold should be used to stratify subtypes for EMT subtype. By default, the MP and EP subtype labels would be determined by the pre-trained Youden index value (recommended), users could change the minPosterior and useMinPosterior to optimize their own results. However, minPosterior functions differently when choosing TCGA subtyping system, it is used to classify samples as Unclassified if probabilities of all four subtypes are below this value.

After predicting the EMP subtypes of GSE62254, the survival outcomes of the predicted labels would be used to check the subtype clinical relevance. To this end, two packages named `survival` and `survminer` should be installed first:

```{r, eval=TRUE}
  if(!requireNamespace('survival')){
    BiocManager::install('survival')
  }

  if(!requireNamespace('survminer')){
    BiocManager::install('survminer')
  }

  library(survival)
  library(survminer)
  GSE62254_subtype[['EMP']] <- emp.res$subtype
  ggsurvplot(
    survfit(Surv(OS.m, Death) ~ EMP, data = GSE62254_subtype),
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5 ,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Overall survival (%)',
    legend.title = '',
    legend.lab = c('EP', 'MP'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254 - OS')
  
  ggsurvplot(
    survfit(Surv(DFS.m, Recur) ~ EMP, data = GSE62254_subtype),
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Disease-free survival (%)',
    legend.title = '',
    legend.lab = c('EP', 'MP'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254 - DFS')
```

## Case study 2: The ACRG subtype prediction

To predict the ACRG subtypes, except changing the `method` parameter from `EMP` to `ACRG`, the rest is the same:

```{r, eval=TRUE, message=TRUE}
  acrg.res <- classifyGC(
    Expr = GSE62254, 
    method = 'ACRG', 
    idType = 'SYMBOL'
  )
  table(acrg.res$subtype, GSE62254_subtype$ACRG.sub)
```

```{r, eval=TRUE}
GSE62254_subtype$ACRG <- acrg.res$subtype
survfit(Surv(OS.m, Death) ~ ACRG, data = GSE62254_subtype) %>%
  ggsurvplot(
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5 ,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Overall survival (%)',
    legend.title = '',
    legend.lab = c('MSI', 'MSS/EMT', 'MSS/TP53-', 'MSS/TP53+'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254')

survfit(Surv(DFS.m, Recur) ~ ACRG, data = GSE62254_subtype) %>%
  ggsurvplot(
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Disease-free survival (%)',
    legend.title = '',
    legend.lab = c('MSI', 'MSS/EMT', 'MSS/TP53-', 'MSS/TP53+'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254')
```

## Case study 3: The TCGA subtype prediction

To predict the TCGA subtypes, set the `method` parameter to `TCGA`, the rest parameters are the same:

```{r, eval=TRUE, message=TRUE}
  tcga.res <- classifyGC(
    Expr = GSE62254,
    method = 'TCGA',
    idType = 'SYMBOL'
  )
```

```{r, eval=TRUE}
GSE62254_subtype$TCGA <- tcga.res$subtype
survfit(Surv(OS.m, Death) ~ TCGA, data = GSE62254_subtype) %>%
  ggsurvplot(
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5 ,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Overall survival (%)',
    legend.title = '',
    legend.lab = c('CIN', 'EBV', 'GS', 'MSI'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254')

survfit(Surv(DFS.m, Recur) ~ TCGA, data = GSE62254_subtype) %>%
  ggsurvplot(
    pval = TRUE,
    conf.int = F,
    risk.table = T,
    risk.table.col = "strata",
    size = 1.5,
    pval.size = 8,
    xlab = 'Follow up (months)',
    ylab = 'Disease-free survival (%)',
    legend.title = '',
    legend.lab = c('CIN', 'EBV', 'GS', 'MSI'),
    ggtheme = theme_classic2(),
    palette = "Dark1"
  ) + ggtitle('GSE62254')
```

# The Shiny application
To facilitate non-R users and simplify programming burdens, a Shiny application of molecular subtype prediction based on gene expression profiles was provided by running function `classifyGC_interface()` and the Shiny application will automatically appear on the default browser (Chrome, Firefox, Edge and Safari recommended). Example dataset could be found on the `Analyze` page and subtype prediction results could be downloaded directly in various export formats. For more information, please run `??classifyGC_interface` to see the details.  

```{r, eval=FALSE}
  classifyGC_interface()
```

Of note, the errors of running the Shiny application might differ due to the complex conditions of uploaded gene expression profiles, and handling all potential errors is impractical. Please try to follow the rules of example dataset, and users should carefully check the requirements of uploading gene expression profiles:

-   Currently, the expression profiles with 'NA' values cannot be uploaded and processed by GCclassifier. It is suggested to perform imputation before uploading to the application.\
-   Please choose the right type of gene identifiers according to your data. Currently, GCclassifier can accept NCBI Entrez, Ensembl, HGNC symbol and Refseq.\
-   Keywords such as "SYMBOL", "ENSEMBL", "ENTREZID" and "REFSEQ" cannot be included in the file columns.   
-   Only numeric values in gene expression profiles are accepted.\
-   Gene expression profiles should not contain any negative value(s).   

If you encounter error information of using Shiny web application, please carefully check your uploaded gene expression profiles, and strictly follow the format of example dataset.   

# Session Info
```{r, echo=FALSE}
  sessionInfo()
```

# References
